Stability of variational quantum dynamics with Gaussian approximants

### 2013-07-26 paper.tex
writing

\def\G{|Γ〉}
\def\cite#1{$[\hbox{\tt #1}]$}

@@

Define a restricted operator
$$|Γ_χ〉f=π^{-½}\int |α〉e^{-½|α|²}χ(α*)f(α*)\,d²α.$$
For now, suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  The most obvious examples are characteristic functions, where $χ$ is 1 inside a circle of radius $b$, and 0 outside.  However, the formalism allows more gradual truncation as well.  We can rewrite this
$$\eqalign{|Γ_χ〉f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
	&=2\sqrt π∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr,}$$
where
$$f(re^{iθ})=∑_{m=-∞}^∞ f_m(r)e^{imθ}.$$
Presumably, the right singular vectors have the $f_n$ of least norm such that 
$$2\sqrt π(n!)^{-½}\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr=1.$$

What is the appropriate norm to use?  In the case that $χ$ is a characteristic function on the domain $D$, it is presumably the truncated Bargmann norm
$$〈g,f〉=\int_D e^{-|α|²}g*(α)f(α)\,d²α=\int e^{-|α|²}χ(α)g*(α)f(α)\,d²α.$$
Note that this is only an inner product when $g$ and $f$ are restricted to $D$.  In this case, we wish to minimise 
$$‖f_n‖²=2π\int_0^b re^{-r²}f²_n(r)\,dr,$$
given 
$$2\sqrt π(n!)^{-½}\int_0^b r^{n+1}e^{-r²}f_n(r)dr=1.$$
If one considers changing $f$ by $Δa$ in an interval around $a$, the Lagrange condition is
$$re^{-r²}f_n(r)\propto r^{n+1}e^{-r²},$$
or 
$$f_n(r)\propto rⁿ.$$
The analytic functions still win.  Let $f_n(r)=rⁿ$.  Then
$$‖f_n‖²=2π\int_0^b r^{2n+1}e^{-r²}\,dr$$
and
$$‖ |Γ_χ〉f_n‖=2\sqrt π(n!)^{-½}\int_0^b r^{2n+1}e^{-r²}dr,$$
so we have
$$σ_n={‖ |Γ_χ〉f_n‖\over ‖f_n‖}=\sqrt{2\over n!}
	\left(\int_0^b r^{2n+1}e^{-r²}dr\right)^½.$$
According to Wolfram,
$$\int_0^b r^{2n+1}e^{-r²}dr=½\left(Γ(n+1)-Γ(n+1,b²)\right),$$
where $Γ$ is Mathematica's version of the incomplete gamma function.  Therefore, recalling that $n!=Γ(n+1)$,
$$σ_n=\left(1-{Γ(n+1,b²)\over Γ(n+1)}\right)^½.$$
Unfortunately, Matlab defines the incomplete gamma function in a different way than Mathematica does.  Fortunately, that Matlab function, {\tt gammainc}, is exactly the square of the right-hand side of the above equation.

@@

\topinsert \XeTeXpicfile gweight.pdf width \hsize 
The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.  The points $x_i$ are normally distributed (top), or evenly spaced (bottom).  Note how nearby points tend to cancel each others' samples.  I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
\endinsert

@@

\centerline{\XeTeXpicfile cats.pdf width \hsize}

\centerline{\XeTeXpicfile catpicard.pdf width \hsize}

\centerline{\XeTeXpicfile catsws.pdf width \hsize}

\centerline{\XeTeXpicfile catl.pdf width \hsize}

@@

%% Singular value decompositions

Variational dynamics seeks an expansion $g$ to approximate a quantum state $|ψ〉≈Gg$.  Given the form of Equation~1, this requires solving a Fredholm integral equation of the first kind.  That is the canonical example of an ill-posed problem: rounding and discretisation errors in $|ψ〉$ are magnified enormously in the solution $g$.  To solve these problems, it helps to understand the singular values and vectors of the operator $G$.

The singular value expansion of the linear operator $G$ has the form
$$Gf = ∑_{j=1}^∞ |g_j〉σ_j〈v_j,f〉,$$
where the left singular vectors $|g_j〉$ are orthonormal kets, the singular values $σ_j$ are positive real numbers, the right singular vectors $v_j$ are normalised entire functions, and presumably $〈,〉$ is the Bargmann inner product.  This can be written
$$G = |G〉SV†,$$
where $|G〉$ is a row vector of the left singular kets, $S$ is a matrix with the singular values along the diagonal, and $V$ maps complex coefficients to linear combinations of the functions $v_j$.

Because the expansions are underdetermined, the set $|g_j〉$ will be complete, but the set $v_j$ will not be.  Alternatively, some of the $v_j$ have zero singular values and no corresponding left singular vector.  One can write a full SVE, where $V†$ is a full expansion, but $S$ truncates the vector of coefficients.

@@

>>>>

\def\G{|Γ〉}
\def\Gx{|Γ_χ〉}
\def\cite#1{$[\hbox{\tt #1}]$}

@@

%% The SVE

The above paragraph can be stated quantitatively.  Define a restricted operator
$$\Gx f=π^{-½}\int |α〉e^{-½|α|²}χ(α*)f(α*)\,d²α.$$
Most obviously, $χ$ could be a characteristic function, with $χ(z)$ being 1 for $|z|≤b$ and 0 outside this disk.  The formalism allows gradual truncation too.

The sizes of states and their expansions can be read from the singular value expansion of $\Gx $.   If the domain and range of $\G$ were the same space, it would have an eigenvalue decomposition; the SVE is a generalisation of that.  It has the form
$$\Gx f = ∑_{i=1}^∞ |u_i〉σ_i〈v_i,f〉.$$
The kets $|g_i〉$ are an orthonormal basis of state space, known as the left singular vectors of $\Gx$.  The $σ_i$ are positive real numbers, called singular values.  The functions $v_i$ are are orthonormal under the Bargmann inner product, and are called right singular vectors.

In the case where $χ=1$, and $\Gx=\G$, the expansion is quite arbitrary.  The left singular kets can be any orthonormal basis, the right singular vector $v_i$ is the unique entire functions for which $\G v_i=|u_i〉$, and every singular value is 1.  If the domain and range of $\G$ were the same, this would indicate that it was a projector.  The component of $|ψ〉$ in the null space of $\G$ is discarded, being orthogonal to every $v_i$, while the other component keeps the same expansion coefficients, while the basis $v_i$ is replaced by the $|u_i〉$.

When the operator is restricted, the SVE gives more information. If $\Gx$ were one to one, each $v_i$ would be the unique preimage of $|g_i〉$; since $\Gx$ is singular, $v_i$ is the preimage with least norm.  The singular values are the ratios of $‖\Gx v_i‖$ to $‖v_i‖$.  If $|u_i〉$ has a low particle number, it can be expanded efficiently: $v_i$ will be close to $〈Γ⁺|u_i〉$, and $σ_i$ will be close to 1.  If $|u_i〉$ includes components with large particle number, its expansion will have a larger norm than itself.  In this case, $v_i$ will be a truncation of $〈Γ⁺|u_i〉$ to small amplitudes, and $σ_i$ will be small.

Before we study the singular values and vectors of $\Gx$, consider why small singular values are a problem.  The pseudoinverse of $\Gx$ with least norm is
$$〈Γ⁺_χ|=∑_{i=1}^∞ v_iσ_i^{-1}〈u_i|.$$  Suppose this is applied to a state $|ψ〉$ that has a low expected particle number, but small components with large particle number.  Such a component will overlap some $|u_i〉$ belonging to small $σ_i$, and $〈Γ⁺_χ|$ will multiply it by $σ_i^{-1}$, generating large components of those $v_i$ in the expansion function.  In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.  When the expansion is discretised, however, it is very easy for the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contains the important information about $|ψ〉$ to disappear in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.

Let's compute some singular values and vectors.  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
$$\eqalign{\Gx f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
	&=2\sqrt π∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr,}$$
where
$$f(re^{iθ})=∑_{m=-∞}^∞ f_m(r)e^{imθ}.$$
Presumably, the right singular vectors have the $f_n$ of least norm such that 
$$2\sqrt π(n!)^{-½}\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr=1.$$

What is the appropriate norm to use?  In the case that $χ$ is a characteristic function on the domain $D$, it is presumably the truncated Bargmann norm
$$〈g,f〉=\int_D e^{-|α|²}g*(α)f(α)\,d²α=\int e^{-|α|²}χ(α)g*(α)f(α)\,d²α.$$
Note that this is only an inner product when $g$ and $f$ are restricted to $D$.  In this case, we wish to minimise 
$$‖f_n‖²=2π\int_0^b re^{-r²}f²_n(r)\,dr,$$
given 
$$2\sqrt π(n!)^{-½}\int_0^b r^{n+1}e^{-r²}f_n(r)dr=1.$$
If one considers changing $f$ by $Δa$ in an interval around $a$, the Lagrange condition is
$$re^{-r²}f_n(r)\propto r^{n+1}e^{-r²},$$
or 
$$f_n(r)\propto rⁿ.$$
The analytic functions still win.  Let $f_n(r)=rⁿ$.  Then
$$‖f_n‖²=2π\int_0^b r^{2n+1}e^{-r²}\,dr$$
and
$$‖ \Gx f_n‖=2\sqrt π(n!)^{-½}\int_0^b r^{2n+1}e^{-r²}dr,$$
so we have
$$σ_n={‖ \Gx f_n‖\over ‖f_n‖}=\sqrt{2\over n!}
	\left(\int_0^b r^{2n+1}e^{-r²}dr\right)^½.$$
According to Wolfram,
$$\int_0^b r^{2n+1}e^{-r²}dr=½\left(Γ(n+1)-Γ(n+1,b²)\right),$$
where $Γ$ is Mathematica's version of the incomplete gamma function.  Therefore, recalling that $n!=Γ(n+1)$,
$$σ_n=\left(1-{Γ(n+1,b²)\over Γ(n+1)}\right)^½.$$
Unfortunately, Matlab defines the incomplete gamma function in a different way than Mathematica does.  Fortunately, that Matlab function, {\tt gammainc}, is exactly the square of the right-hand side of the above equation.

@@

\topinsert \XeTeXpicfile gweight.pdf width \hsize 
The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.  The points $x_i$ are normally distributed (top), or evenly spaced (bottom).  Note how nearby points tend to cancel each others' samples.  I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.

The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.  This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.  Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.  Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.  When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.  This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
\endinsert

@@

@@

@@

%% ill-posedness

As shown above, the restricted expansion operator $\Gx $ has singular values that decay to very small values, whose singular vectors are states with particle numbers corresponding to coherent amplitudes outside the support of $χ$.

\topinsert\XeTeXpicfile cats.pdf width \hsize\endinsert

\topinsert\XeTeXpicfile catpicard.pdf width \hsize\endinsert

\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert

\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
